---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

```{r}
file <- unzip("activity.zip")
origDF <- read.csv(file, header = TRUE, sep = ",", na.strings = "NA", nrows = 17568)


imputedDF <- origDF
noNADF <- na.omit(origDF)

```

## What is mean total number of steps taken per day?

```{r 1.Histogram }
library(sqldf)
totSteps <- sqldf("select sum(steps) from noNADF group by date ")
colnames(totSteps) <- "steps"

hist(totSteps$steps , breaks = 25, ylim = c(0,10), col = 'brown', xlab = 'Sum of Steps', bg = 'white', main = 'Number of Steps per day')

```

## What is the mean and medium

```{r 2.Mean and Median}
mmPDate <- sqldf("select date as Date, AVG(steps) as Mean, ((max(steps) + min(steps)) / 2)  as Median from noNADF group by date ")


```
### Report of the Mean and Median
```{r }

print(mmPDate)

```

## What is the average daily activity pattern?

```{r Average Daily Activity Pattern}

library(lattice)

aveStepsInt <- sqldf("select interval, AVG(steps) as Mean from noNADF group by interval ")
 
xyplot(aveStepsInt$Mean ~ aveStepsInt$interval,
        data = aveStepsInt,
        type = "l", lty = c(1, 2, 2, 1),
        xlab = "Interval", ylab = "Average Steps"
       )
 
```

## What is the interval at maximum number of steps

```{r Interval at Maximum number of steps}

maxInterval <- sqldf(" select interval from noNADF where steps = (select MAX(steps) from noNADF)")
printStr <- sprintf("The 5 min interval with maximum number of steps = %d",maxInterval$interval)
print(printStr)

```

## Imputing missing values

```{r Impute}

totalNA <- nrow(origDF) - nrow(noNADF)
printStr <- sprintf("Total number of Missing Values = %d",totalNA)
print(printStr)

# Imputing the NAs with the mean for that interval from the No NAs data frame
for(i in 1:17568) {
        if (is.na(origDF$steps[i])) { 
            imputedDF$steps[i] <- aveStepsInt$Mean[aveStepsInt$interval 
                                             == origDF$interval[i]]
        }
}
colnames(imputedDF) <- c("steps", "date", "interval")

```

## Histogram for imputed values

```{r 2.Histogram }

totSteps2 <- sqldf("select sum(steps) from imputedDF group by date ")
colnames(totSteps2) <- "steps"

hist(totSteps2$steps , breaks = 25, ylim = c(0,10), col = 'brown', xlab = 'Sum of Steps', bg = 'white', main = 'Number of Steps per day')

```

## Mean and medium for imputed values

```{r Mean and Median}
mmPDate2 <- sqldf("select date as Date, AVG(steps) as Mean, ((max(steps) + min(steps)) / 2)  as Median from imputedDF group by date ")

mmPDate2$Median <- as.integer(mmPDate2$Median)

```
### Report of the Mean and Median for imputed values
```{r }

print(mmPDate2)

```

## Do these values differ from the estimates from the first part of the assignment?

###Comparing the mean/median between the datasets with the missing values removed and the missing values replaced by mean, they look identical. 


## What is the impact of imputing missing data on the estimates of the total daily number of steps?

### There does not seem to be any effect of imputing the NAs with MEAN since the overall mean remains the same and the median is not affected anyway.

### However the 8 days which had all missing values earlier and did not show up in the first report now show up.



## Are there differences in activity patterns between weekdays and weekends?

```{r Differences in activity patters between weekdays and weekends}

#create a vector of weekdays
imputedDF$day <- weekdays(as.Date(imputedDF$date))
weekdays1 <- c('Monday', 'Tuesday', 'Wednesday', 'Thursday', 'Friday')
#Use `%in%` and `weekdays` to create a logical vector
#convert to `factor` and specify the `levels/labels`
imputedDF$wDay <-  factor((imputedDF$day %in% weekdays1)+1L,
      levels=1:2, labels=c('weekday', 'weekend'))

```

## Making a time series panel plot for weekdays versus weekends

```{r time series panel plot}

aveStepsInt2 <- sqldf("select interval,
                      wDay,
                      AVG(steps) as Mean 
                      from imputedDF 
                      where wDay = 'weekday'
                      group by interval ")
aveStepsInt3 <- sqldf("select interval,
                      wDay,
                      AVG(steps) as Mean 
                      from imputedDF 
                      where wDay = 'weekend'
                      group by interval ")


df4 <- rbind(aveStepsInt3, aveStepsInt2)


xyplot(Mean ~ interval | wDay, 
        data = df4,
        layout=c(1,2),
        type = "l", lty = c(1, 2, 2, 1),
        xlab = "Interval", ylab = "Number of Steps"
       )


```


